\documentclass[10pt,twocolumn]{article}
\usepackage[T1]{fontenc}
\usepackage[utf8]{inputenc}
\usepackage{lmodern}
\usepackage[margin=1.8cm,top=2.4cm,bottom=2cm,columnsep=0.6cm]{geometry}
\usepackage{fancyhdr}
\usepackage{titlesec}
\usepackage{booktabs}
\usepackage{amsmath,amssymb,amsfonts}
\usepackage{microtype}
\usepackage{xcolor}
\usepackage{mdframed}
\usepackage{parskip}
\usepackage{enumitem}
\usepackage{hyperref}

\definecolor{rulered}{RGB}{180,30,30}
\definecolor{boxgray}{RGB}{245,245,245}
\definecolor{keywordgray}{RGB}{100,100,100}

\pagestyle{fancy}
\fancyhf{}
\fancyhead[L]{\textsc{\small Claude Mag}}
\fancyhead[C]{\small\textit{Machine Learning Research Digest}}
\fancyhead[R]{\small 2026-07-22}
\fancyfoot[C]{\small\thepage}
\renewcommand{\headrulewidth}{0.8pt}

\titleformat{\section}{\normalsize\bfseries\color{rulered}}{}{0pt}{}%
  [\vspace{-4pt}\color{rulered}\rule{\columnwidth}{0.4pt}]
\titlespacing{\section}{0pt}{8pt}{4pt}

\hypersetup{colorlinks=true,urlcolor=rulered,linkcolor=black}

\begin{document}
\setlength{\parindent}{0pt}
\setlength{\parskip}{4pt}

{\small\color{keywordgray}\textbf{Keywords:}
  speculative decoding \textbullet{}
  mixture-of-experts \textbullet{}
  inference efficiency \textbullet{}
  expert routing \textbullet{}
  LLM serving}\par\vspace{4pt}

{\LARGE\bfseries\raggedright
  EcoSpec}\par\vspace{4pt}

{\large\itshape\raggedright
  Beijing Institute of Technology Identifies Expert Scattering as a
  Fundamental Barrier to Speculative Decoding in MoE Models and Achieves
  Up to 1.62$\times$ Speedup Through Cost-Aware Draft Selection}\par\vspace{6pt}

{\small\textbf{By} Jincheng Xie, Runheng Liu, Heyan Huang, Yawen Ling,
  Hanbin Dai, Yu Zheng, Wen Hu
  (Beijing Institute of Technology)}\par\vspace{2pt}

{\small\color{keywordgray}arXiv:2607.12696 \textbullet{}
  July 2026}
\par\vspace{2pt}\noindent{\color{rulered}\rule{\columnwidth}{0.6pt}}\vspace{6pt}

Speculative decoding accelerates autoregressive generation by drafting
multiple tokens quickly and verifying them in a single forward pass
through the large target model.
For dense models this works cleanly: the verification pass has
a fixed memory footprint.
But for Mixture-of-Experts models --- which power all current
frontier open-source systems including DeepSeek-V3.1 and Qwen3 ---
researchers from Beijing Institute of Technology have identified
a previously uncharacterized failure mode: \emph{expert scattering}.
Their remedy, EcoSpec, jointly optimizes acceptance likelihood and
expert reuse, delivering up to 1.62$\times$ end-to-end speedup
over standard speculative decoding at no quality cost.

\begin{mdframed}[backgroundcolor=boxgray,linecolor=rulered,linewidth=1pt,
    innerleftmargin=6pt,innerrightmargin=6pt,
    innertopmargin=6pt,innerbottommargin=6pt]
\textbf{\small AT A GLANCE}\par\vspace{4pt}
\begin{description}[leftmargin=0pt,labelsep=4pt,itemsep=2pt,parsep=0pt,
    font=\small\bfseries]
  \item[Problem:] Confidence-driven draft selection in MoE models routes
    high-probability tokens to disjoint expert sets, increasing
    expert-weight memory traffic and eliminating the speedup from
    speculation.
  \item[Why it matters:] All major frontier open-source models
    (DeepSeek-V3.1, Qwen3-235B) use MoE; speculative decoding must
    account for expert memory locality to be effective at scale.
  \item[Method:] EcoSpec: lightweight expert predictor + dynamic expert
    buffer; composite draft scoring balancing acceptance likelihood
    and expert reuse fraction.
  \item[Result:] Up to 1.62$\times$ end-to-end speedup on DeepSeek-V3.1
    (671B) across reasoning, coding, QA, and dialogue benchmarks.
  \item[Evidence:] Standard speculative decoding achieves negative
    speedup at draft length $\geq$8 on DeepSeek-V3.1; EcoSpec
    reaches 1.62$\times$ at the same draft length.
\end{description}
\end{mdframed}

\section{The Big Picture}

Speculative decoding has become the standard technique for accelerating
large language model inference: a small draft model generates a sequence
of candidate tokens at low cost, and the large target model verifies
the entire sequence in a single forward pass, accepting tokens that
agree with its own distribution and rejecting the rest.
For dense models, each forward pass has a fixed compute and memory
cost, and the speedup is determined solely by the acceptance rate
of the draft tokens.

MoE models break this assumption.
In an MoE model, each token is routed to a small subset of experts
(typically 2 out of 64), and the memory footprint of a forward pass
is proportional to the number of \emph{unique} experts activated
across all tokens in the batch.
When draft tokens are selected to maximize acceptance likelihood
alone, those tokens may route to many different expert sets ---
forcing the serving system to load additional expert weight matrices
into GPU high-bandwidth memory for each verification step.
This expert scattering can make speculative decoding slower than
standard autoregressive generation.

\section{Why Existing Approaches Fall Short}

Prior speculative decoding methods for MoE models adopt the same
draft selection criterion as for dense models: the draft token with
highest likelihood under the draft model is selected at each position.
This ignores the routing structure of the target model entirely.

Two prior attempts at MoE-aware speculative decoding --- MoESD
and Utility-Driven Speculative Decoding --- optimize the expert
activation pattern for the draft model but not for the target model,
leaving the expert scattering problem at the target verification
pass unresolved.

\section{Core Mechanism}

EcoSpec modifies draft selection by introducing two components.
The \textbf{expert predictor} is a lightweight auxiliary network
that, given the draft model's hidden state for a candidate token,
forecasts which of the target model's 64 experts that token will
activate.
The \textbf{dynamic expert buffer} tracks which expert weights are
currently cached in GPU high-bandwidth memory for the active
verification batch.
Draft selection then uses a composite score that balances the
probability of token acceptance with the fraction of required
experts already in the buffer, so the verification step can
proceed with minimal additional expert loading.

\section{Technical Formulation}

Let $c_i$ be a candidate token at draft position $i$ and let
$p_{\text{accept},i}$ be its acceptance probability under the target
model.
The expert predictor $h_\psi$ maps the draft hidden state to a
binary vector over the expert set:
\[
\hat{e}_i = h_\psi\bigl(d_i\bigr) \in \{0,1\}^E
\]
where $d_i$ is the draft model's hidden state at position $i$
and $E$ is the total number of experts.
The expert reuse fraction is:
\[
f_{\text{reuse},i}
  = \frac{\lvert \hat{e}_i \cap \mathcal{B} \rvert}{\lvert \hat{e}_i \rvert}
\]
where $\mathcal{B}$ is the current expert buffer (the set of
experts in GPU HBM).
The composite draft score is:
\[
s_i = p_{\text{accept},i} \cdot \bigl(1 + \lambda \cdot f_{\text{reuse},i}\bigr)
\]
with $\lambda = 0.4$ chosen by grid search.
After each verification step, $\mathcal{B}$ is updated to include
the experts activated by the accepted tokens.

\section{Learning Procedure}

\begin{enumerate}[itemsep=2pt,parsep=0pt]
  \item \textbf{Expert predictor training:} $h_\psi$ is trained
    offline on a 10K-sample calibration corpus to minimize
    prediction error for target expert activation patterns.
    Training takes approximately 30 minutes on a single GPU and
    adds $<$0.3\% latency to each draft step.
  \item \textbf{Draft model selection:} Any standard speculative
    decoding draft model can be used; EcoSpec modifies only the
    token selection step, not the draft model itself.
  \item \textbf{Buffer initialization:} At the start of each
    generation, the buffer $\mathcal{B}$ is initialized with
    the experts activated by the prompt tokens' verification pass.
  \item \textbf{No target model modification:} EcoSpec requires
    no changes to target model weights, architecture, or serving
    infrastructure beyond the expert predictor and buffer tracking.
\end{enumerate}

\section{Experimental Findings}

\begin{center}
\small
\begin{tabular}{llcc}
\toprule
Model & Task & Std.\ SD & EcoSpec \\
\midrule
DeepSeek-V3.1 & Reasoning & 1.00$\times$ & \textbf{1.62}$\times$ \\
DeepSeek-V3.1 & Coding & 1.00$\times$ & \textbf{1.51}$\times$ \\
Qwen3-235B-A22B & QA & 1.00$\times$ & \textbf{1.44}$\times$ \\
GPT-OSS-120B & Dialogue & 1.00$\times$ & \textbf{1.38}$\times$ \\
\bottomrule
\end{tabular}
\end{center}

Standard speculative decoding achieves negative speedup (slower than
autoregressive) on DeepSeek-V3.1 at draft lengths of 8 or more
tokens, confirming expert scattering as a practical barrier.
EcoSpec reduces the active expert footprint by 18--35\% relative to
confidence-only draft selection while maintaining or improving
token acceptance rates.

\section{Ablations and Interpretation}

\textbf{Expert predictor quality:} Using a perfect oracle predictor
improves speedup to 1.81$\times$ on DeepSeek-V3.1, indicating that
the expert predictor is the current bottleneck and that improving
its accuracy is the most promising direction for future gains.

\textbf{Dynamic vs.\ static buffer:} A static expert buffer (fixed
set, not updated per step) achieves only 1.22$\times$ speedup ---
less than half the gain of the dynamic variant --- confirming
that tracking live HBM state is essential for effective cost-aware
draft selection.

\textbf{Lambda sensitivity:} Speedup on DeepSeek-V3.1 varies from
1.41$\times$ ($\lambda = 0$, pure confidence) to 1.62$\times$
($\lambda = 0.4$) to 1.48$\times$ ($\lambda = 1.0$, pure reuse).
The optimal $\lambda$ is consistent across models and tasks,
suggesting that a moderate balance between confidence and
reuse is universally beneficial.

\textbf{Acceptance rate:} EcoSpec's composite scoring reduces the
acceptance rate by at most 1.2 percentage points relative to
confidence-only selection on any tested benchmark, confirming
that the speedup comes from reduced expert loading rather than
a higher acceptance rate --- and that the quality of generated
text is unchanged.

\section*{Reference}

Jincheng Xie, Runheng Liu, Heyan Huang, Yawen Ling,
Hanbin Dai, Yu Zheng, Wen Hu.\
\textbf{Less Experts, Faster Decoding: Cost-Aware Speculative
Decoding for Mixture-of-Experts}.
arXiv:2607.12696, July 2026.
\url{https://arxiv.org/abs/2607.12696}

\end{document}
